---
title: "Text clustering"
author: 
  - Osaro Orebor 1168827
  - Máté Csikós-Nagy 4395565
  - Fani Profiti 1240390
  - Yara Yachnyk 4913329
  - Niels Wagenaar 3133998
date: last-modified
format:
  html:
    toc: true
    self-contained: true
    code-fold: true
    df-print: kable
execute:
  warning: false
  message: false
  echo: true
---

```{r}
#| label: R packages
#| echo: false
#| warning: false
#| message: false

library(text2vec)
library(wordcloud)
library(magrittr)
library(tm)

# additional packages here
```

```{r}
#| label: data loading

data("movie_review")

# your R code to load the data here
```

# Data description

Describe the data and use a visualization to support your story. (approx. one or two paragraphs

```{r}
#| label: eda visualization
#| warning: false

# your R code to generate the plot here
movie_review %$% wordcloud(review, 
                           min.freq = 10, 
                           max.words = 50, 
                           random.order = FALSE,
                           colors = brewer.pal(8, "Dark2"))

```

# Text pre-processing

Describe text pre-processing steps you have used (approx. one or two paragraphs)

# Text representaion

Briefly describe your text representation method. (approx. one or two paragraphs)

# Text clustering

Briefly describe which models you compare to perform clustering. (approx. two or three paragraphs)

# Evaluation & model comparison

Describe how you compare the methods and why. (approx. two or three paragraphs)


```{r}
#| label: table example
data.frame(
  model       = c("clustering model 1", "clustering model 2"),
  performance = c(1.2, 1.8),
  other       = c(0.5, 0.3),
  notes       = c("Some note", "another note")
)
```


# Team member contributions

Write down what each team member contributed to the project.

- Author One: a, b, c
- Author Two: b, c, d
- Author Three: a, b, d
